---
title: "Exploring Inhalation Toxicity of Chemical Compounds"
author: "Dylann Yavorski, Norah Nguyen"
date: "05/13/2025"
output:
  pdf_document:
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE, message = FALSE, results = FALSE, warning=FALSE}
# ALL LIBRARY STATEMENTS HERE
library(NbClust)
library(mclust)
library(flexclust)
library(cluster)
library(fpc)
library(knitr)
library(tidyverse)
library(dplyr)
library(readr)
library(ggplot2)
library(ROSE)
library(caret)
library(ROCR)

# these declarations are a good way to ensure that all random numbers in your report use the same settings:
rand.v <- "4.3.2"
seed.val <- 123456
# see line 60 for how they are referenced.
```

# Introduction

> Humans are constantly exposed to a diverse array of synthetic and naturally occurring chemical compounds in their environment. Many of these substances are incorporated into everyday consumer products, including fragrances. Therefore, it is critical to determine whether the chemical compounds used in fragrance formulations are truly safe for human health, particularly with respect to inhalation exposure.\
Our study on fragrance toxicity involved three datasets: one containing chemical compounds known to be toxic when inhaled, another consisting of fragrance compounds with unknown toxicity, and a third matching the unknown fragrance compounds with an associated toxicity level-low (Class I), intermediate (Class II), and high (Class III). In the third dataset, toxicity outcomes were obtained using the Toxtree module's Revised Cramer Rules [1], a tool available for free download and widely used in toxicology for predicting toxicity based on chemical structure.\
In our analysis, compounds classified as Class II or III were labeled as toxic, while Class I compounds were considered safe. We sourced these datasets from PubChem, a chemical database maintained by the National Institutes of Health (NIH) [2], and the Environmental Protection Agency (EPA) [3]. Initially, we obtained a list of chemicals from PubChem and then cross-referenced it with the EPA database to gather a comprehensive dataset that includes the relevant variables.\
The dataset of toxic compounds totals roughly 1500, while the unknown fragrance and toxtree datasets total roughly 3000. After some cleaning in Excel, the unknown fragrance dataset observations were reduced to about 1600. In our study, we shuffled the unknown and toxtree data to create a merged mixed toxic and non-toxic compound dataset for regression analysis.\
We have two central questions regarding our datasets:\ 
1. Are Ames Test Predictions significant in identifying similarities between toxic compounds? 
2. What variables regarding toxic and non-toxic chemical compounds are the strongest predictors for inhalation toxicity?\
For our first modeling method, we employed clustering to explore the toxic inhalation dataset in relation to Ames Test Predictions. In chemistry, the Ames Test is used to discern the capacity for a chemical to trigger cell mutations (mutagenesis), which may lead to cancer in humans. A positive Ames test value indicates higher mutagenicity, while a negative test value indicates no genotoxicity. We chose to use this variable as it was present in the toxic dataset and not the unknown, so we wished to understand if Ames Test results were at all meaningful in finding similarities between toxic compounds.\
For our second modeling method, the variables we chose for further study through regression in the fragrance compound and toxic inhalation datasets included vapor pressure, water solubility, and logP, as these properties are correlated with compound inhalation toxicity and were shared variables in both datasets. We aimed to identify which of these predictors most strongly influence inhalation toxicity, with the potential to refine the model and incorporate other relevant predictors in future work.

# Data Processing

> While reading in all datasets, we used the na.strings function to identify rows containing N/A in the datasets and convert them to NA's understood by R. Next, we removed extraneous columns irrelevant to our study in the toxic dataset and changed categorical columns (such as names of compounds and compound identifiers) to factors in all datasets. After, we performed a right-join merge on the toxtree data and unknown fragrance compound dataset to create a new known dataset, while removing duplicate values. This reduced the new dataset observations to 63. Then, we removed NA values and extreme outliers for some variables in the toxic dataset, though as we are dealing with chemical compounds a wide variance in values still remained. Removing outliers and missing values for the toxic dataset reduced the total observations to 387. Finally, we performed an outer-join merge on the toxic dataset and the new, mixed dataset, removing duplicate rows shared between the two to use later in our regression model. This merged dataset has 422 observations.

```{r echo=FALSE}
# This is where you write R code that reads in the data, cleans it, and assigns any categorical types to factors.
inhalation1.df <- read.csv('fragrance_chemicals_EPA.csv', na.strings = "N/A")
inhalation2.df <- read.csv('Inhalation_risks_EPA.csv', na.strings = "N/A")
inhalation_toxic.df <- read.csv('post_toxtree_fragrance.csv', na.strings = "N/A")


# removing extra columns and columns with high NA count that will not be helpful as a result
inhalation2.df <- inhalation2.df[, -53:-780]
inhalation2.df <- select(inhalation2.df, -25:-27)
inhalation2.df <- select(inhalation2.df, -46)
inhalation2.df <- select(inhalation2.df, -39)
inhalation2.df <- select(inhalation2.df, -38)

# categorical to factors
inhalation1.df$PREFERRED_NAME <- factor(inhalation1.df$PREFERRED_NAME)
inhalation1.df$X.Compound.CID <- factor(inhalation1.df$X.Compound.CID)
inhalation1.df$SMILES <- factor(inhalation1.df$SMILES)
inhalation1.df$Synonyms <- factor(inhalation1.df$Synonyms)
inhalation1.df$INCHI_STRING <- factor(inhalation1.df$INCHI_STRING)
inhalation1.df$InChIKey <- factor(inhalation1.df$InChIKey)
inhalation1.df$IUPAC.Name <- factor(inhalation1.df$IUPAC.Name)

inhalation1.df$MOLECULAR_FORMULA <- factor(inhalation1.df$MOLECULAR_FORMULA)

inhalation2.df$PREFERRED_NAME <- factor(inhalation2.df$PREFERRED_NAME)
inhalation2.df$FOUND_BY <- factor(inhalation2.df$FOUND_BY)
inhalation2.df$SMILES <- factor(inhalation2.df$SMILES)
inhalation2.df$INPUT <- factor(inhalation2.df$INPUT)
inhalation2.df$INCHI_STRING <- factor(inhalation2.df$INCHI_STRING)
inhalation2.df$MOLECULAR_FORMULA <- factor(inhalation2.df$MOLECULAR_FORMULA)
inhalation2.df$DTXSID <- factor(inhalation2.df$DTXSID)
inhalation2.df$QSAR_READY_SMILES <- factor(inhalation2.df$QSAR_READY_SMILES)
inhalation2.df$SAFETY_DATA <- factor(inhalation2.df$SAFETY_DATA)
inhalation2.df$EXPOCAST <- factor(inhalation2.df$EXPOCAST)
inhalation2.df$DATA_SOURCES <- factor(inhalation2.df$DATA_SOURCES)
inhalation2.df$CASRN <- factor(inhalation2.df$CASRN)
inhalation2.df$TOXCAST_NUMBER_OF_ASSAYS.TOTAL <- factor(inhalation2.df$TOXCAST_NUMBER_OF_ASSAYS.TOTAL)
inhalation2.df$TOXICITY <- factor(inhalation2.df$TOXICITY)

inhalation_toxic.df$CASRN <- factor(inhalation_toxic.df$CASRN)
inhalation_toxic.df$PREFERRED_NAME <- factor(inhalation_toxic.df$PREFERRED_NAME)
inhalation_toxic.df$TOXICITY <- factor(inhalation_toxic.df$TOXICITY)
inhalation_toxic.df$SMILES <- factor(inhalation_toxic.df$SMILES)

# merging toxtree results with data points in unknown dataset
fragrance.df <- merge(inhalation_toxic.df, inhalation1.df, by = "SMILES", all.y = TRUE)
fragrance.df <- unique(fragrance.df)
colnames(fragrance.df)[6] = "PREFERRED_NAME"
# removing extra preferred name column
fragrance.df <- select(fragrance.df, -3)

# handling NA and outliers
fragrance.df <- na.omit(fragrance.df)
inhalation2.df <- na.omit(inhalation2.df)
inhalation2.df <- inhalation2.df[inhalation2.df$VAPOR_PRESSURE_MMHG_TEST_PRED < 400,]
inhalation2.df <- inhalation2.df[inhalation2.df$AMES_MUTAGENICITY_TEST_PRED > 0,]
inhalation2.df <- inhalation2.df[inhalation2.df$AMES_MUTAGENICITY_TEST_PRED < 1,]

# merged dataset for regression
merged_toxic.df <- merge(fragrance.df, inhalation2.df, by = c("PREFERRED_NAME", "SMILES", "TOXICITY", "INCHI_STRING", "MOLECULAR_FORMULA", "AVERAGE_MASS","MONOISOTOPIC_MASS","CASRN","VAPOR_PRESSURE_MMHG_TEST_PRED", "WATER_SOLUBILITY_MOL.L_TEST_PRED", "XLOG_P"), all = TRUE)
merged_toxic.df <- select(merged_toxic.df, 1:11)
merged_toxic.df <- merged_toxic.df[-c(4,6,14,18,21,62,67,76,107,120,125,203,218,221,246,243,267,271,283,318,326,338,344,355,361,362,363,390 ),]
#NOTE: ALL R code will be hidden by the echo=FALSE chunk attribute so that the reader does not see it. 
```

# Data Exploration

> First, we analyzed the distribution of Ames test predictions in the toxic dataset. The histogram plot and summary table below for the Ames Mutagenicity Test Predictions show that the data is right-skewed, with the largest portion of observations having a test prediction between 0-0.2. This indicates that the majority of observations in the toxic chemicals dataset have slight mutagenetic properties.

\newpage

```{r echo=FALSE, out.width="80%", out.height="80%", fig.align = 'center', fig.cap="Distribution of Ames Mutagenesis Test Plot"}
# your plot code will go here.
hist(inhalation2.df$AMES_MUTAGENICITY_TEST_PRED, breaks = 7, main = "Distribution of Ames Mutagenesis Test", xlab = "Test Prediction Value", col = "blue")

```


```{r echo=FALSE, fig.keep='all', fig.cap="Distribution of Ames Mutagenesis Test Table"}

# use the kable function for all tables-make sure to add the caption as it will take care of the title and uniquely number all tables in your report.

ames_test.tbl <- inhalation2.df %>%
    summarise(Min = min(AMES_MUTAGENICITY_TEST_PRED),
              Max = max(AMES_MUTAGENICITY_TEST_PRED),
              Median = median(AMES_MUTAGENICITY_TEST_PRED),
              Mean = mean(AMES_MUTAGENICITY_TEST_PRED))
kable(ames_test.tbl, caption = "Descriptive Statistics for Ames Mutagenicity Test Predictions:")

```
\newpage

# Method 1: Clustering

> In this section, we aim to discover more about the toxic dataset through two clustering methods: K-Means and Gaussian Mixture. We hope to learn if there are any meaningful groupings in our toxic dataset, and if these groupings are at all correlated with Ames Test Prediction values.

# Training and Testing Data Sets

> To create proper clusters, we first removed all categorical variables from the toxic dataset, assigning the result to a new dataframe, inhalation2.data. Then, we scaled the results as there is wide variety between predictor values. To determine ideal parameters for clustering, used the NbClust package to automate the ideal kmeans cluster amount. The results below indicate an efficient cluster number would be around 6-8, where the graph comes to a "point," or "elbow."

```{r echo=FALSE, results='hide', fig.keep='all', fig.cap = "NbClust Algorithm For Determining Cluster Size: "}
#Note these statements use the variables declared in the library load chunk at the beginning.
RNGversion(rand.v)
set.seed(seed.val)

# code to make up the training and testing sets here.
# eliminates character/factor columns from data set and assigned to new df variable
inhalation2.data <- select_if(inhalation2.df, is.numeric)




inhalation2.data.scaled <- scale(inhalation2.data[-1])
nc <- NbClust(inhalation2.data.scaled, min.nc=2, max.nc=15, method="kmeans")
 kable(table(nc$Best.n[1,]), caption = "NbClust Results")
 
```
\newpage

# Modeling

> In this section, we employed two different clustering methods- K-Means clustering, and Gaussian Mixture Model-Based clustering. Our K-Means model is fed a pre-determined amount of clusters, or groups, and identifies data points' locations based on the value of each cluster's center, or centroid. As for Gaussian Mixture clustering, the model also aims to label data as being part of a particular cluster, but does this using probability distributions as opposed to means. As for a variable of interest, we chose the Ames Mutagenicity test for its smaller distribution compared to several other potential toxicity predictors with wider variance.

## K-Means Clustering

> To fit a k-means cluster model, we passed in the scaled inhalation data, instructing the model to use 7 clusters. We decided to use a low nstart value of 5 due to the large variance in the dataset even after scaling, which distributed the data between clusters more meaningfully.

```{r echo=FALSE}
# code to fit the model
RNGversion(rand.v)
set.seed(seed.val)

fit.inhalation <- kmeans(inhalation2.data.scaled, 7, nstart= 5)
```

### K-Means Model Fit Results

> The size of each cluster varies widely, with the smallest cluster having only 10 values, and the largest cluster with 126. The measure of distance within clusters is greater than the measure of distance between clusters, meaning there is likely no significant trend among values in the dataset as a whole.

```{r echo=FALSE,, fig.keep='all', fig.cap = "Inhalation Fit Model Parameters: "}
# code to fit the model

inhalation_cluster <- table(fit.inhalation$size)
kable(inhalation_cluster, col.names = c("Cluster Amount", "Cluster"), caption = "Size Of Each Cluster")

inhalation_cluster <- table(fit.inhalation$tot.withinss)
kable(inhalation_cluster, col.names = c("Measure", "Freq"), caption = "Measure of Within Distance Of Each Cluster")

inhalation_cluster <- table(fit.inhalation$betweenss)
kable(inhalation_cluster, col.names = c("Measure", "Freq"),caption = "Measure of Between Distance Of Each Cluster")
```
\newpage

### K-Means Clustering Prediction Performance

> We then analyzed the fit.inhalation model in terms of one of our variables of interest, Ames Mutagenicity Test Prediction. The table below analyzes the scaled means for each cluster, while the plot illustrates the clustering model itself. The table and plot appear to reflect the data above- there are certainly enough differences between values as a whole to create clusters, but there is still greater variety within clusters than between them. This may indicate the data itself is generally similar- as these are all toxic compounds with some degree of mutagenicity.

```{r echo=FALSE, out.width="50%", out.height="50%", fig.align = 'center', fig.cap="Ames Test Cluster Means"}
# code to evaluate the model
Ames_Test <- fit.inhalation$centers[, 10]
Ames.Col = c(Ames_Test)
Ames.Test.df <- data.frame(
  Means = Ames.Col, 
  Cluster_Number = c(1,2,3,4,5,6,7)
)
kable(Ames.Test.df, caption = "Means For Each Cluster of Ames Test")
```

```{r echo=FALSE, out.width="80%", out.height="80%", fig.align = 'center', fig.cap="K-Means Clustering For Ames Test"}
# your plot code will go here.

plotcluster(inhalation2.data.scaled, fit.inhalation$cluster, xlab = "Cluster Range X", ylab = "Cluster Range Y")
```

\newpage

## Gaussian Mixture Model-Based Clustering- Mclust

> To fit the Mclust Gaussian Mixture model, we passed in the scaled inhalation data with no other parameters, as this method uses a mixture of models.

```{r echo=FALSE}
# code to fit the model
RNGversion(rand.v)
set.seed(seed.val)

inhalation.model.clus <- Mclust(inhalation2.data.scaled)
mclust.inhalation <- summary(inhalation.model.clus)
```

### Mclust Fit Results

> The Mclust model compared 14 different fitted models, with VEV being the most accurate until about 6 clusters. After that point, the EEV model is more accurate in determining cluster belonging.

```{r echo=FALSE,, fig.keep='all', fig.cap = "Mclust Inhalation Fit Model Parameters: "}
# code to fit the model

plot(inhalation.model.clus, inhalation2.data.scaled, what="BIC")

```

\newpage


### Mclust Clustering Prediction Performance

> We then analyzed the inhalation.model.clus model in terms of our variable of interest, Ames Mutagenicity Test Prediction. We used the chi-square test to measure correlation between the chosen clusters and Ames Test Predictions. The returned p-value of 0.025 indicates there is a significant amount of correlation between Ames Test Pred values of data points and their associated clusters.


```{r echo=FALSE, message=FALSE, warning=FALSE, out.width="50%", out.height="50%", fig.align = 'center', fig.cap= "Chi Square Test for Mclust and Ames Test"}
# code to evaluate the model
model.clus.tbl <- chisq.test(table(inhalation.model.clus$classification, inhalation2.df$AMES_MUTAGENICITY_TEST_PRED))


Chisq_t <- model.clus.tbl
Chisq.col = c(Chisq_t$statistic, Chisq_t$parameter, Chisq_t$p.value, Chisq_t$data.name)
Chisq.Test.df <- data.frame(
  Parameter = c("X-Squared: ","df: ","p-value: ","Data Name: "),
  Data_Value = Chisq.col
)
kable(Chisq.Test.df, caption= "Pearson's Chi-squared Test:")
```

> The plot below is a cleaner visual representation using dimension reduction of the ideal cluster belonging returned by the Mclust model. We can see through both K-Means and Gaussian Mixture Modeling that most of the clusters themselves are relatively similar to one another, but there may be one significantly different grouping of data in the dataset, indicated by cluster 7 in the K-Means Test and the yellow squares in the dimensional modeling below. The chi-square test results with a significant p-value also show that the Ames Test Predictions column could be useful to further categorize and understand the chemical properties of the given compounds.

```{r echo=FALSE, out.width="80%", out.height="80%", fig.align = 'center', fig.cap="Dimensional Reduction Cluster Plot for Mclust Model"}
# your plot code will go here.

inhalation.model.dr <- MclustDR(inhalation.model.clus)
plot(inhalation.model.dr, what = "contour")
```

\newpage

# Method 2: Logistic Regression Classification Model

> For our second model, we employ logistic regression. Using our merged dataset created earlier, and the toxicity variable, our goal is to predict a binary outcome: “toxic” or “non-toxic.” We will evaluate the model’s performance based on its accuracy and misclassification rate, and identify which predictors are most influential in determining inhalation toxicity. First, we analyzed the frequency of toxic and non-toxic observations in our merged dataset. As shown below in the barplot and frequency table, roughly ~11% of the data is in the minority class (safe), making our dataset moderately imbalanced.

```{r echo=FALSE, out.width="80%", out.height="80%", fig.align = 'center', fig.cap="Frequency of Toxicity Plot"}
# your plot code will go here.
ggplot(merged_toxic.df, aes(x=reorder(TOXICITY, TOXICITY, function(x)-length(x)))) + geom_bar(fill="red") + labs(x="Toxicity", y="Count")
```

```{r echo=FALSE, fig.keep='all', fig.cap="Toxicity Frequency Table"}

# use the kable function for all tables-make sure to add the caption as it will take care of the title and uniquely number all tables in your report.

toxicity.tbl <- table(merged_toxic.df$TOXICITY)
kable(toxicity.tbl, col.names = c("Toxicity", "Frequency"), caption = "Frequency Levels of Toxicity Observations")
```

\newpage

# Training and Testing Data Sets

> We opted to shuffle our data points in the merged dataset, creating a new dataframe (shuffled_merge.df) before partitioning training and testing sets, as we merged both datasets together and there are far more toxic than non-toxic observations. This allows for the most balanced data possible in each set. We then created training and testing sets, using a standard 75% of data for the training set and the remaining 25% for testing. Finally, we used the ROSE function to further balance the training set before fitting a model.


```{r echo=FALSE, results='hide', fig.keep='all'}
#Note these statements use the variables declared in the library load chunk at the beginning.
RNGversion(rand.v)
set.seed(seed.val)

# create randomly shuffled dataset
shuffled_merge.df <- merged_toxic.df[sample(1:422),]
# code to make up the training and testing sets here.
shuffled_merge.df <- droplevels(shuffled_merge.df)
index.train <- createDataPartition(y = shuffled_merge.df$TOXICITY, p = 0.75, list = FALSE)
train.data <- shuffled_merge.df[index.train,]
test.data <- shuffled_merge.df[-index.train,]

rose.data <- ROSE(TOXICITY ~ ., data=train.data, seed=1)$data
 
```

# Modeling

> In this section, we fit a logistic regression model to our training and testing data to determine which predictors (XLog_P, Water Solubility, Vapor Pressure) were significant in identifying compound toxicity. Logistic regression is a method for determining a binary outcome using independent variables (predictors), and is an ideal model type as our outcome variable is binary in nature, either toxic or not.

## Logistic Regression Model

> We decided to fit the model to our three predictors of interest, XLog_P, Water Solubility, and Vapor Pressure. We used the glm method with ten cross-validation folds.

```{r echo=FALSE, results='hide', fig.keep='all'}
# code to fit the model
RNGversion(rand.v)
set.seed(seed.val)

reg.mod <- train(form = TOXICITY ~ XLOG_P + WATER_SOLUBILITY_MOL.L_TEST_PRED + VAPOR_PRESSURE_MMHG_TEST_PRED, data = train.data,
  trControl = trainControl(method="cv", number =10),
  method = "glm", family = "binomial")

```
\newpage

### Logistic Regression Model Fit Results

> Our fitted model result table includes serveral different calculations. The "intercept" refers to the constant term in the dataset when all predictors are set to zero, while standard error and z values measure the variance of predicted outcomes using a predictor variable compared to the true value. The estimate parameter compares the projected change in our outcome, toxicity, if each predictor were to be increased by one unit. Finally, the p-value measures the statistical significance of each predictor value on the outcome, with values less than 0.05 deemed to be significant. In our case, the model determined that XLog_P is mildly significant in determining toxicity levels, while water solubility and vapor pressure were less important.\
We then decided to plot an ROC curve, which measures prediction accuracy in a model. The larger the AUC (Area Under the Curve), the better the model. As shown below, our fitted toxicity model has an AUC measure of 0.57, which is only marginally better than a random model with a value of 0.50.


\newpage

```{r echo=FALSE, fig.keep='all', fig.cap = "GLM Logistic Regression Model Results: "}
# code to fit the model
reg.mod.df <- data.frame(
  Coefficients = c("(Intercept)", "XLOG_P", "WATER_SOLUBILITY_MOL.L_TEST_PRED", "VAPOR_PRESSURE_MMHG_TEST_PRED"),
  Estimate=c(2.2605357, -0.2059483, 0.0421158, -0.0002361),
  Std_Error=c(0.2959516, 0.1008555, 0.1862724, 0.0009452),
  Z_Value=c(7.638, -2.042, 0.226, -0.250),
  P_Value=c(2.2e-14, 0.0412, 0.8211, 0.8027))
kable(reg.mod.df, caption = "Logistic Regression Model Predictor Results")


```

```{r echo=FALSE, out.width="80%", out.height="80%", fig.align = 'center', fig.cap="Toxicity Model Accuracy"}
# your plot code will go here.
mod.probs <- predict(reg.mod, newdata = test.data, type = "prob")[,2]
# get the predictions with reference to the true values in test set.
pred.vals <-prediction(mod.probs,test.data$TOXICITY)
#ROC plot
# The performance function allows us to get many metrics for evaluation.
# Type ?performance in the console and look at the Details section in the document page for a listing.
perf.val <- performance(pred.vals, 'tpr', 'fpr')
plot(perf.val, col = 'red', lwd = 1.5)
abline(0,1,lty=3) # add the diagonal line
#auc- using the perfromance function again with'auc' this time.
perf.mod<- performance(pred.vals,'auc',fpr.stop=1)
auc.mod<-slot(perf.mod,"y.values")[[1]]
legend("bottomright", legend=c(paste("AUC = ",round(auc.mod, 2))), cex=0.8)

```
\newpage

### Logistic Regression Prediction Performance

> To further understand the performance of our model, we created a confusion matrix table to compare true positive and true negative rates with false positive and false negative rates. The confusion matrix results in the table below indicate the model was able to correctly identify toxic compounds, but was unable to identify safe compounds, labeling them toxic as well. We believe this is due to the imbalanced nature of the dataset and hence a lack of non-toxic observations to train our model on.\
We then created a plot of the sensitivity (the proportion of identified true positives out of all actual positive values) and the specificity (the proportion of identified true negatives out of all actual negative values). This plot identifies the best threshold of values for determining the outcome classification. In this case, the plot determined that the ideal threshold is about 0.89. However, due to the lack of available training data for "safe" predictions, we were unable to test our predictions at such a high threshold.

```{r echo=FALSE, out.width="50%", out.height="50%", fig.align = 'center', fig.cap= "Confusion Matrix Results"}
# code to evaluate the model
mod.preds<-predict(reg.mod, newdata = test.data)

matrix.results <- confusionMatrix(mod.preds, test.data$TOXICITY, positive='Toxic')

matrix.df <- data.frame(Reference = c("Safe", "Toxic"),
                        Safe_Prediction = c(0, 12),
                        Toxic_Prediction = c(0,93)
)
kable(matrix.df, caption="Predicted versus Actual Compound Toxicity Results")
```

```{r echo=FALSE, out.width="80%", out.height="80%", fig.align = 'center', fig.cap="Sensitivity/Specificity Plot for Toxicity Levels in The Dataset"}
# your plot code will go here.
perf.vals.sens <- performance(pred.vals, 'sens')
perf.vals.spec <- performance(pred.vals, 'spec')
# A variable for the default threshold- added to this plot for comparison.
default.thresh <- 0.50
#plot the two curves
plot(perf.vals.spec, , xlab="Threshold", col = 'black', lwd = 1.5)
plot(perf.vals.sens, add=TRUE, col = 'red', lwd = 1.5)
# add labels to the series
mtext("Sensitivity",side=4, padj=-2, col='red')
# get threshold at the point where sensitivity = specificity
sens = cbind(unlist(performance(pred.vals, "sens")@x.values), unlist(performance(pred.vals, "sens")@y.values))
spec = cbind(unlist(performance(pred.vals, "spec")@x.values), unlist(performance(pred.vals, "spec")@y.values))
best.thresh <- sens[which.min(apply(sens, 1, function(x) min(colSums(abs(t(spec) - x))))), 1]
abline(v = best.thresh, col = "blue", lty = 4)  
abline(v = default.thresh, col = "grey", lty = 4) 
legend("bottomright", legend=c(paste("best thresh=",round(best.thresh, digits=3)), paste("default thresh= ",default.thresh)),
  )

```

\newpage

# Conclusions

> To address our first research question—whether Ames Test predictions are significant in identifying similarities between toxic compounds—we applied two clustering techniques: K-Means and Gaussian Mixture Models. While both methods produced at least one distinct cluster, most clusters did not show meaningful differentiation, and there was no strong separation based on mutagenicity. Although Ames Test predictions showed limited statistical significance, they did not align consistently with the overall cluster groupings. This suggests that Ames Test results are not strong indicators of structural or feature-based similarity among toxic compounds as captured by unsupervised clustering methods.\
To explore our second research question—identifying the strongest predictors for inhalation toxicity—we applied logistic regression to a merged dataset of toxic and non-toxic compounds. Among the variables analyzed, XLog_P (a measure of lipophilicity) was the only statistically significant predictor. However, the model's predictive performance was only marginally better than random, likely due to class imbalance and the limited availability of complete non-toxic observations.\
Key challenges included managing missing data, interpreting overlapping clusters, and selecting biologically meaningful features. Future work should address class imbalance by incorporating more complete non-toxic cases and explore additional predictor variables supported by toxicological literature to improve model robustness. Additionally, exploring alternative modeling strategies such as ensemble methods could move beyond binary classification and better capture the complex, continuous nature of toxicological effects.

# References

> 1. Toxtree module's Revised Cramer Rules:\
https://toxtree.sourceforge.net/cramer.html \
2. PubCHEM Datasets:\
https://pubchem.ncbi.nlm.nih.gov/classification/#hid=72&search=Fragrances \
https://pubchem.ncbi.nlm.nih.gov/classification/#hid=72&search=Inhalation+Risk \
3. EPA Datasets:\
https://comptox.epa.gov/dashboard/batch-search